---
title: "Flex RT with R"
output: html_notebook
---

The FWX file will be export to Microsoft Excel. Which can be edited easily. Then exported back to FWX.
```{r}
if (! "xlsx" %in% rownames(installed.packages())) {
  # install package to manipulate Excel files
  install.packages("xlsx")
}
library('xlsx')
```

Bit operations like shift left
```{r}
library("bitops")
```

Reads the main configuration file: Binary strings, images and settings.
```{r}
fwx.name <- "G:/Documents and Settings/WinXP/My Documents/simatic/PDATA.FWX"
fwx.size <- file.size(fwx.name)
fwx.handle <- file(fwx.name, "rb")
fwx.data <- readBin(fwx.handle, "raw", fwx.size)
# clean up
close(fwx.handle)
rm("fwx.handle")
```

Functions:
```{r}
# Read 16 bits unsigned little endian word
d2 <- function(pos) {
  arr2 <- fwx.data[(pos+1):(pos+2)]
  d <- as.integer(arr2)
  d[1] + d[2] * 256
}

# Read 32 bits little endian double word
d4 <- function(pos) {
  arr4 <- fwx.data[(pos+1):(pos+4)]
  d <- as.integer(arr4)
  d[1] + bitShiftL(d[2], 8) + bitShiftL(d[3], 16) + bitShiftL(d[4], 24)
}
```

Not sure if it should be 16bit or not, because it always looks the same, the data type are always Little Endian (LSB first)
The first 6 words of 16bit are always the same
```{r}
header.start1 <- d2(0)
header.start2 <- d2(2)
header.start3 <- d2(4)
header.start4 <- d2(6)
header.start5 <- d2(8)
header.start6 <- d2(0xa)
```

These two double words seems to say how big is the padding at the last section of the file
It looks like it should be in a 4 bytes alignment (e.g. 0xe2 is actually 0xe4 bytes)
```{r}
header.post_tables_padding1 <- d4(0xc)
header.post_tables_padding2 <- d4(0x10)
```

This is the end of the tables, this way we can tell what is the length of the last object.
```{r}
header.tables_end <- d4(0x14)
```

Most important part, the Table Of Contents (TOC), or the array of arrays. This part tells in which offset every part of the file is. It does not say the length of each part, but you can calculate it by subtracting the next offset from the current.
```{r}
# how many entries in the table of contents
header.toc_entries <- d4(0x18)
# location of table of content, should be the last part in the file
header.toc_offset <- d4(0x1c)
```

I believe this tells the run time which elements to initialize
```{r}
header.init_entries <- d4(0x20)
header.init_offset <- d4(0x24)
```

Just made up some names for these parts. These are 16 bits (double bytes) entries.
```{r}
header.info_entries <- d4(0x28)
header.info_offset <- d4(0x2c)
header.metainfo_entries <- d4(0x30)
header.metainfo_offset <- d4(0x34)
```

How many languages exist in the file and what is their code pages (0x409 - English, 0x407 - German, 0x40D - Hebrew)
```{r}
header.lang_entries <- d4(0x38)
header.lang_offset <- d4(0x3c)
```

Read and store the Table Of Contents (array of arrays)
Entries with 0 are empty
Entries with non-zero contain the offset in the file of the begining of each table
```{r}
# initalize empty TOC array
toc <- array(0,header.toc_entries)
# fill in the TOC with offsets of all arrays in the file
for(i in 1:header.toc_entries) {
  toc[i] <- d4(header.toc_offset + (i-1)*4)
}
```

Read the names of all the tables (arrays) in the file
```{r}
# create an empty data frame for the toc entries
toc.entries <- data.frame()
# iterate thru the TOC to get all table names
for(i in 1:length(toc)) {
  offset <- toc[i]
  if (offset != 0) {
    # how many entries exist in the current table
    entries <- d2(offset)
    # table type? all tables have this type of 0x64
    type <- d2(offset+2)
    # parent id? all tables have this 0x01 number
    parent <- d2(offset+4)
    # table id - this number is identical to the index i
    id <- d2(offset+6)
    # name length - in 16bits UTF16 Little Endian characters
    len <- d2(offset+8)
    # name of table
    name_start <- offset+0xa+1
    name_characters <- list(fwx.data[(name_start):(name_start+len*2-1)])
    name <- iconv(name_characters, "UTF-16LE", "UTF-8")
    
    # create a data.frame row
    next_pos <- dim(toc.entries)[1]+1
    toc.entries[next_pos, 'entries'] <- entries
    toc.entries[next_pos, 'type'] <- type
    toc.entries[next_pos, 'parent'] <- parent
    toc.entries[next_pos, 'id'] <- id
    toc.entries[next_pos, 'name'] <- name
    toc.entries[next_pos, 'offset'] <- offset
    # save the item size
    if (next_pos > 1) {
      toc.entries[next_pos-1, 'item_size'] <- offset - last_offset
    }
    last_offset <- offset
  }
}
# calculate the last item size with the global constant
toc.entries[next_pos, 'item_size'] <- header.tables_end - last_offset
```

Read the Tags, this is the list of all variables connected to the server monitoring the system.
```{r}
# create empty data frame to hold all tags
tags.entries <- data.frame()

tags_index <- toc.entries$name == 'VAR'
tags <- toc.entries[tags_index,]
# location of offsets of all objects in this table has a constant length
tags_offsets <- tags$offset + 0x34
# iterate thru all the VARs (tags) and get their names
for(i in 1:tags$entries) {
  # location of offset of current item
  tag_offset <- tags_offsets + (i-1)*4
  # relative offset of current item (from beginning of datablock)
  tag_offset_relative <- d4(tag_offset)
  # absolute offset of current item (from beginning of file)
  tag_offset <- tag_offset_relative + tags_offsets + tags$entries*4

  # String type? Either 3 or 0xb (which is 3+8 - probably some flag)
  type <- d2(tag_offset)
  # name length - in 16bits UTF16 Little Endian characters
  len <- d2(tag_offset+2)
  # name of tag
  name_start <- tag_offset+4 + 1
  name_characters <- list(fwx.data[(name_start):(name_start+len*2-1)])
  name <- iconv(name_characters, "UTF-16LE", "UTF-8")
  # must align by 4 bytes the length of the string (in bytes)
  align <- (len*2) %% 4
  data_offset <- tag_offset+4+len*2+align
  # Some data?
  data_len <- d4(data_offset) # length of current tag, almost always = 2
  data53 <- d2(data_offset+4) # this value is almost always 0x35=53
  idx <- d2(data_offset+6) # this is a running index if data53==53
  dat_offset <- d2(data_offset+8) # this is almost always zero except in cases where data53 != 53
  data80E3 <- d4(data_offset+10) # this value is always = 0x80E3 
  data00 <- d2(data_offset+14) # this value is always = 0
  
  # create a data.frame row
  next_pos <- dim(tags.entries)[1]+1
  tags.entries[next_pos, 'type'] <- type
  tags.entries[next_pos, 'name'] <- name
  tags.entries[next_pos, 'offset'] <- tag_offset
  tags.entries[next_pos, 'data_len'] <- data_len
  tags.entries[next_pos, 'data53'] <- data53
  tags.entries[next_pos, 'idx'] <- idx
  tags.entries[next_pos, 'data_offset'] <- dat_offset
  tags.entries[next_pos, 'data80E3'] <- data80E3
  tags.entries[next_pos, 'data00'] <- data00
  
  if (data_len > 0) {
    data01 <- d2(data_offset+16) # always 0
    data227 <- d2(data_offset+18) # almost always 227
    # data type can be:
    # 0x82: INT - signed 16 bits word between -32768 and +32767
    # 0x83: LONG - signed 32 bits double word between -2147483648 and +2147483647
    # 0x84: Real - between -3.402823E+38 and +3.402823E+38
    # 0x85: Double - between -1.79769313486232E+308 and +1.79769313486232E+308
    # 0x86: VT_CY
    # 0x87: VT_DATE
    # 0x88: STRING
    # 0x89: VT_DISPATCH
    # 0x8a: VT_ERROR
    # 0x8b: BOOL - Boolean value of either 0 or -1
    # 0x90: Char - between -128 and 127
    # 0x91: BYTE - Unsigned 8 bits values between 0 and 255
    # 0x92: UINT - Unsigned 16 bits values between 0 and 65535
    # 0x93: ULong - Unsigned 32 bits vlues between 0 and 4294967295
    # 0x94: VT_I8
    # 0x95: VT_UI8
    # Arrays begin with 0x2000
    # 0x2002: Arrays of INTs
    # 0x2092: Arrays of UINTs
    data_type <- d2(data_offset+20)
    
    array_size <- d2(data_offset+22) # number of sub element, usually = 1 unless it is an array
    
    data_pos <- d2(data_offset+24) # successive numbers with skips of almost always 4
    data02 <- d2(data_offset+26) # this value is always = 0
    data03 <- d2(data_offset+28) # this value is always = 0
    data04 <- d2(data_offset+30) # this value is always = 0
    
    # for data_len 3 - everything shifted by 8 bytes
    if (data_len == 3) {
      data227 <- data02
      data_type <- data03
      array_size <- data04
      data_pos <- d2(data_offset+32)
      data02 <- d2(data_offset+34) # this value is always = 0
      data03 <- d2(data_offset+36) # this value is always = 0
      data04 <- d2(data_offset+38) # this value is always = 0
    }

    tags.entries[next_pos, 'data01'] <- data01
    tags.entries[next_pos, 'data227'] <- data227
    tags.entries[next_pos, 'data_type'] <- data_type
    tags.entries[next_pos, 'array_size'] <- array_size
    tags.entries[next_pos, 'data_pos'] <- data_pos
    tags.entries[next_pos, 'data02'] <- data02
    tags.entries[next_pos, 'data03'] <- data03
    tags.entries[next_pos, 'data04'] <- data04
  }
  
  # save the item size
  if (next_pos > 1) {
    tags.entries[next_pos-1, 'item_size'] <- tag_offset - last_offset
  }
  last_offset <- tag_offset
}
# calculate the last item size with the size of the parent table
tags.entries[next_pos, 'item_size'] <- tags$offset + tags$item_size - last_offset
# View the tags (TODO: Create HTML to view all tags)
View(tags.entries)
```
