---
title: "Flex RT with R"
output: html_notebook
---

The FWX file will be export to Microsoft Excel. Which can be edited easily. Then exported back to FWX.
```{r}
if (! "xlsx" %in% rownames(installed.packages())) {
  # install package to manipulate Excel files
  install.packages("xlsx")
}
library('xlsx')
```

Bit operations like shift left
```{r}
library("bitops")
```

Reads the main configuration file: Binary strings, images and settings.
```{r}
fwx.name <- "G:/Documents and Settings/WinXP/My Documents/simatic/PDATA.FWX"
fwx.size <- file.size(fwx.name)
fwx.handle <- file(fwx.name, "rb")
fwx.data <- readBin(fwx.handle, "raw", fwx.size)
# clean up
close(fwx.handle)
rm("fwx.handle")
```

Functions:
```{r}
# Read 16 bits unsigned little endian word
d2 <- function(pos) {
  arr2 <- fwx.data[(pos+1):(pos+2)]
  d <- as.integer(arr2)
  d[1] + d[2] * 256
}

# Read 32 bits little endian double word
d4 <- function(pos) {
  arr4 <- fwx.data[(pos+1):(pos+4)]
  d <- as.integer(arr4)
  d[1] + bitShiftL(d[2], 8) + bitShiftL(d[3], 16) + bitShiftL(d[4], 24)
}
```

Not sure if it should be 16bit or not, because it always looks the same, the data type are always Little Endian (LSB first)
The first 6 words of 16bit are always the same
```{r}
header.start1 <- d2(0)
header.start2 <- d2(2)
header.start3 <- d2(4)
header.start4 <- d2(6)
header.start5 <- d2(8)
header.start6 <- d2(0xa)
```

These two double words seems to say how big is the padding at the last section of the file
It looks like it should be in a 4 bytes alignment (e.g. 0xe2 is actually 0xe4 bytes)
```{r}
header.post_tables_padding1 <- d4(0xc)
header.post_tables_padding2 <- d4(0x10)
```

This is the end of the tables, this way we can tell what is the length of the last object.
```{r}
header.tables_end <- d4(0x14)
```

Most important part, the Table Of Contents (TOC), or the array of arrays. This part tells in which offset every part of the file is. It does not say the length of each part, but you can calculate it by subtracting the next offset from the current.
```{r}
# how many entries in the table of contents
header.toc_entries <- d4(0x18)
# location of table of content, should be the last part in the file
header.toc_offset <- d4(0x1c)
```

I believe this tells the run time which elements to initialize
```{r}
header.init_entries <- d4(0x20)
header.init_offset <- d4(0x24)
```

Just made up some names for these parts. These are 16 bits (double bytes) entries.
```{r}
header.info_entries <- d4(0x28)
header.info_offset <- d4(0x2c)
header.metainfo_entries <- d4(0x30)
header.metainfo_offset <- d4(0x34)
```

How many languages exist in the file and what is their code pages (0x409 - English, 0x407 - German, 0x40D - Hebrew)
```{r}
header.lang_entries <- d4(0x38)
header.lang_offset <- d4(0x3c)
```

Read and store the Table Of Contents (array of arrays)
Entries with 0 are empty
Entries with non-zero contain the offset in the file of the begining of each table
```{r}
# initalize empty TOC array
toc <- array(0,header.toc_entries)
# fill in the TOC with offsets of all arrays in the file
for(i in 1:header.toc_entries) {
  toc[i] <- d4(header.toc_offset + (i-1)*4)
}
```

Read the names of all the tables (arrays) in the file
```{r}
# create an empty data frame for the toc entries
toc.entries <- data.frame()
# iterate thru the TOC to get all table names
for(i in 1:length(toc)) {
  offset <- toc[i]
  if (offset != 0) {
    # how many entries exist in the current table
    entries <- d2(offset)
    # table type? all tables have this type of 0x64
    type <- d2(offset+2)
    # parent id? all tables have this 0x01 number
    parent <- d2(offset+4)
    # table id - this number is identical to the index i
    id <- d2(offset+6)
    # name length - in 16bits UTF16 Little Endian characters
    len <- d2(offset+8)
    # name of table
    name_start <- offset+0xa+1
    name_characters <- list(fwx.data[(name_start):(name_start+len*2-1)])
    name <- iconv(name_characters, "UTF-16LE", "UTF-8")
    
    # create a data.frame row
    next_pos <- dim(toc.entries)[1]+1
    toc.entries[next_pos, 'entries'] <- entries
    toc.entries[next_pos, 'type'] <- type
    toc.entries[next_pos, 'parent'] <- parent
    toc.entries[next_pos, 'id'] <- id
    toc.entries[next_pos, 'name'] <- name
    toc.entries[next_pos, 'offset'] <- offset
  }
}
toc.entries
```
